---
title: "02 - Calculation of Indices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{02 - Calculation of Indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sf)
library(mapme.vegetation)
```

In this tutorial we are explaining how vegetation indices are calculated under the
hood of the package. The most important tool we are using is the [gdalcubes](https://cran.r-project.org/web/packages/gdalcubes/index.html)
package. This packages makes it very easy to create spatio-temporal aggregates
of satellite observations for a given area of interest (AOI) and user defined
aggregation methods and time-frames. Here, we assume that you already downloaded
a number of files using the process outlined in the [Download](E01-download.html)
article with the option `create_cloudmask` enabled. The function will not work
on Sentinel 2 directories which do not contain cloud masks. However the cloud mask
can be created after the download by calling `createCloudMask()` on the safe 
directories.

In order to check the available indices you can use the `check_indices()` function
which will deliver you a data.frame with available indices. We thankfully adapted 
these indices from the awesome`sen2r::list_indices()` package. Here we print the first 10 available 
indices.

```{r check-indices}
head(check_indices(),10)
```

To calculate an index we will have to specify its short name found in the `name`
column of the data.frame. Currently, there is support for a total number of 206
different indices, however, the function can only process one index at a time thus
if you are interested in more than one index you should loop over the function call.

Here is what the function call to calculate the NDVI looks like assuming that
in the files variables the output of a recursive file search for `.jp2` and `.tif` 
files on the SAFE directories is found.

```{r calc, eval = F}
files = list.files("<path>/<to>/<SAFES>", pattern = ".jp2|.tif", recursive = T, full.names = TRUE)
aoi = st_read(system.file("extdata", "testregion.gpkg", package = "mapme.vegetation"))
aoi = st_transform(aoi, st_crs("EPSG:3857"))
rasterfiles = calcIndex(files,
                        aoi = aoi, 
                        epsg = "EPSG:3857", 
                        dx = 20, 
                        dy = 20, 
                        dt = "P1M", 
                        aggregation = "mean", 
                        resampling = "bilinear", 
                        tmpagg = FALSE,
                        timeframe = "complete", 
                        index = "NDVI", 
                        label = "NDVI", 
                        outdir = ".", 
                        ncores = 2)
```

Let's go through the parameters one by one. The `files` variable was already explained above.
It comprises all single band files **including** the cloud mask for each Sentinel 2 image.
Note that the default file names cannot be changed because `calcIndex()` retrives the
associated timestemp for each file from the actual filename. The aoi represents
an `sf` object to determine the spatial extent of the analysis by computing the 
bounding box of the object and using its edge coordinates. You actually do not have
to transform the object to the desired output projection beforehand because the 
function will do so internally if the crs differs, however, it is good advise to 
manually transform the `aoi`. In the `epsg` variable we define the desired output
reference system. This can be any EPSG code, proj4 string or other formats which are
understood by GDAL (see `gdalcubes::cube_view()` for more information). 
`dx` and `dy` are the size of the resulting pixels in x and y direction. Their unit
depends on the reference system sepecified in the `epsg` option, thus extra care
should be taken that to get things right here. In the example above we are using the
Web Mercator projection thus the unit is represented as 20 meters. `dt` specifies
the resolution of the time dimension as an ISO8601 string, thus in the example
above it represents a monthly resolution. This means that after the spatial warping
to the desired CRS and spatial resolution, all images of the same month will be aggregated.
The method of the temporal aggregation is specifed in the `aggregation` option while
any GDAL supported image transformation for the spatial warping can be specified in
the `resampling` option. 

In the example above we set `tmpagg = FALSE` which means
that the time dimension is not going to be reduced in the end of the calculation.
In the present case this is the desired behaviour because we are interested in 
monthly NDVI values. However, let us assume that we downloaded a seasonal dataset
comprising the all Sentinel 2 datasets for the months June and July for each of 
the years 2016 to 2019. Our goal is to retrive one raster for each year indicating
the mean between these two months. Unfortunately, `gdalcubes` does not support irregular 
time dimensions, for this reason we would have to process each year individually. 
What we do not want in this case is to aggregate the raw reflectance values over 
the complete two months and then calculate the NDVI. Rather we want the NDVI calculated
individually for each image, aggregate the images to a monthly resolution and then
calculate the mean of the NDVI for the two months. In cases like this we can set
`tmpagg = TRUE` which will reduce the resulting time dimension by applying the mean 
calculation. In these cases we also would have to set `timeframe = "seasonal"`
to let the function know to calculate on different years seperatly. If this option
is specified, you also have to specify a `years` vector containing the different years
for the calculation as characters. 

The `outdir` and `label` parameters expect charachter vectors. The first specifying the
directory where the output files are written and the second specyfing a label which is
appended to the filename. Finally the parameters `threads` expects a numeric value
indicating how many threads should be made available to the calculation for parallel
processing.

The function will return a character vector pointing to the written `GTiff` files.
The number of output files depends on how the function treats the time dimension because
one file is written per final time dimension. The naming convention follows: `<outdir>/<label><timestemp>.tif`

The next step in a given analysis now would be to extract zonal statistics for
a number of polygons we are interested in from the raster files we processed here.
This is explained in the next article on [zonal statistics](E03-zonalstats.html).



