---
title: "Cloud Masking"
date: "Last modified: `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cloud Masking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette functionality of the `mapme.vegetation` package to enrich the 
sen2cor cloud mask is presented. The issue with the sen2cor scene cover classification
(SCL) is that cloud shadows and pixels at the edge of clouds are not really well 
detected. When this is not accounted for, the quality of subsequent analysis
can be seriously limited. `mapme.vegetation` supplies users with a fast routine
based on GRASS GIS to select pixel which shall be masked and apply a spatial buffer
to enrich the original SCL classification. As a prerequisite to use this
functionality it is assumed that a working GRASS installation is found on your 
machine. We will use exemplary files which have been reduced in their resolution
in order to be able to ship them with the package.

```{r setup, eval = F}
library(mapme.vegetation)
s2_files = list.files(system.file("extdata/s2a", package = "mapme.vegetation"), ".tif", full.names = T)
scl_files = s2_files[grep("SCL", s2_files)]
rundir = file.path(tempdir(), "mapme.vegetation")
dir.create(rundir, showWarnings = F)

scl_buffer(scl_files = scl_files, 
           mask_values = c(1,2,3,7,8,9,10,11), # which values to mask?
           mask_buffer = 500, # distance of buffers (in meters), here exceptionally large to show differences
           grass_bin = "/usr/bin/grass78", 
           threads = 1, 
           outdir = rundir)
```

Let' us compare the original SCL scene to the buffered version. We will use 
the `terra` package for a quick visualization.

```{r show = F}
s2_files = list.files(system.file("extdata/s2a", package = "mapme.vegetation"), ".tif", full.names = T)
scl_files = s2_files[grep("SCL", s2_files)]
```


```{r}
library(terra)
scl_buffered = rast(list.files(system.file("extdata/scl", package = "mapme.vegetation"), ".tif", full.names = T)[1])
scl_original = rast(scl_files[1])

scl = c(scl_original, scl_buffered)
names(scl) = c("original", "buffered")
plot(scl)
```

On the left hand side we see the original SCL scene. Several values are present 
ranging from 2 to 10. On the right hand side we see the buffered version. In yellow
all the masked pixels are shown. White areas indicate pixels which were not in the
buffer zone of the mask values. All the yellow pixels have a value of 1, the first 
value specified in the vector `mask_value` even though the original raster does not
contain any cells with a value of 1. In summary we get a buffered version of the SCL
scene but we loose additional information on the classes. This should be kept in mind
when using the routine and the value for masking should be selected accordingly.
