---
title: "Download"
date: "Last modified: `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Download}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


The `mapme.vegetation` package allows us to easily download and work with Sentinel-2 (S2) Level 2A datasets.
These datasets represent bottom of atmosphere (BOA) reflectance values processed
with [sen2cor](https://step.esa.int/main/snap-supported-plugins/sen2cor/). 
The data is made available via the [Earth Search](https://stacindex.org/catalogs/earth-search#/) 
STAC API from AWS. `mapme.vegetation` can be used to query the server for a given
spatio-temporal extent of interest and download matching results. 
S2 download from AWS is free of cost and does not require a user account. However,
users should be aware that there is a maximum of 500 items per query. In case
the spatio-temporal extent will lead to more items it is up to the user to 
split up the query into suitable chunks. The package allows for querying the data
without a direct download in order to check the items returned.

```{r }
library(mapme.vegetation)
library(sf)

aoi = st_read(system.file("extdata", "exp_region_wgs.gpkg", package = "mapme.vegetation"), quiet = T)
bbox = st_bbox(aoi)

items = download_aws(bbox = bbox, 
                      after = "2017-05-01", 
                      before = "2017-07-31", 
                      timeframe = "full", 
                      max.cloud = 40, 
                      query_only = TRUE)


```

The query resulted in 11 items matching the spatio-temporal extent with a maximum cloud
cover of 40%.
We can proceed downloading the data by setting `query_only = FALSE` and supply additional
arguments of where to write the files


```{r eval=FALSE}
items = download_aws(bbox = bbox, 
                     after = "2017-05-01", 
                     before = "2017-07-31", 
                     timeframe = "full", 
                     max.cloud = 40, 
                     outdir = "../inst/extdata/s2a", 
                     query_only = FALSE)
```

The above call with download the data sequentially. As a alternative `mapme.vegetation`
allows to download the data via the [aria2](https://aria2.github.io/) download utility.
User's have to install the program manually on their machine and indicate the
path to the aria2 binary. Additionall arguments are required to specify how many
connections to the server should be established to download matching items in parallel.
The function call below will download 4 items in parallel with a total of 4 connections
per file to further speed up the download. Also, the retry argument can be set to
specify the number of time the the function will retry to download items for which
the download failed.

```{r eval=FALSE}
items = download_aws(bbox = bbox, 
                     after = "2017-05-01", 
                     before = "2017-07-30", 
                     timeframe = "full", 
                     max.cloud = 40, 
                     outdir = "../inst/extdata/s2a/", 
                     query_only = FALSE,
                     use_aria = T,
                     aria_bin = "/usr/bin/aria2c", 
                     max_concurrent = 4, 
                     max_connections = 4,
                     retry = 1)
```


Note that in the above function calls `timeframe = "full"`. This way, the function
will download all the matching items between the `after` and `before` dates (inclusive).
If you want to request only specific months but across years specify `timeframe = "seasonal"`
to loop over the years and download the matching time period. Note that the
items are returned as a list with each object representing one of the queried years.

```{r}
items = download_aws(bbox = bbox, 
                      after = "2017-05-01", 
                      before = "2020-07-30", 
                      timeframe = "seasonal", 
                      max.cloud = 40, 
                      query_only = TRUE)
```

If you are only interested in specific bands you can reduce the files to download
by specyfing the assets you wish to download.

```{r eval = F}

items = download_aws(bbox = bbox, 
                      after = "2017-05-01", 
                      before = "2017-07-30", 
                      timeframe = "full", 
                      assets = c("B02", "B04", "B08", "SCL"),
                      max.cloud = 40, 
                      query_only = F)

```


