---
title: "Calculate Indices"
date: "Last modified: `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculate Indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Setup 

```{r setup, warnings = FALSE}
library(mapme.vegetation)
library(sf)
library(terra)
rundir = file.path(tempdir(), "mapme.vegetation")
dir.create(rundir, showWarnings = F)
```


## Available Indices

In this vignette `mapme.vegetation` functionality to derive a dense time series
of vegetation indices is presented. The package comes with a list of over 100
indices that can be calculated for Sentinel 2. Issue the following command
to inspect available indices.


```{r}
head(check_indices())
```

## Calculate Indices

In order to specify one or multiple indices refer to the short name found in the
column `name`. As a prerequisite for calculating indices we assume that S2 data
for a given spatio-temporal AOI has been downloaded as explained in the vignette xx.
Additionally, we assume that the SCL has been enriched using the functionality
explained in the vignette xx. It is however possible to skip the cloud mask
buffering and proceed with the original SCL files.
First, let's read in the necessary files and declare some arguments for the
index calculation.

```{r}
s2_files = list.files(system.file("extdata/s2a", package = "mapme.vegetation"), ".tif", full.names = T)
scl_files = list.files(system.file("extdata/scl", package = "mapme.vegetation"), ".tif", full.names = T)
aoi = st_read(system.file("extdata", "exp_region_wgs.gpkg", package = "mapme.vegetation"), quiet = T)
aoi_utm = st_transform(aoi, st_crs(32638))
bbox = st_bbox(aoi_utm)

dx = 200 # desired outcome resolution in x direction
dy = 200 # desired outcome resolution in y direction
dt = "P14D" # desired temporal resolution - here 1 month
srs = "EPSG:32638" # desired CRS of the outcome raster - bbox must be specified in that crs
after = "2017-05-01" # start date
before = "2017-07-31" # end date
index = "NDVI" # index to calculate, can specify multiple indices
aggregation = "max" # aggregation method for scene within the same time step
resampling = "cubic" # GDAL resampling method to match the outcome grid size
dir.create(file.path(rundir, "ndvi")) # create output directories
dir.create(file.path(rundir, "filled")) # create output directories
label = "NDVI-" # appended to filename
mask_layer = "SCL" # which band in the data cube is used as mask
mask_values = 1 # which values to mask out
mask_invert = F # logical if mask_values should be masked (TRUE) or kept (FALSE)
```


With this prerequisites we can nearly start to calculate the NDVI. Note, however,
that we set the `mask_value = 1` to account for our enriched cloud mask. That is
why we should exchange the original SCL files with our new SCL files.

```{r}
s2_files = s2_files[-grep("SCL", s2_files)]
s2_files = c(s2_files, scl_files)
```

Now we can start calculating the index. We will use a collection definition which
comes with the `mapme.vegetation` package to let `gdalcubes` how to interpret the
filenames from AWS. That way, we are sure that the spatio-temporal extent of
each single file is rightly interpreted.

```{r}
raw_ndvi = index_cube(files = s2_files, 
                      format = system.file("sentinel2_l2a_cog.json", package = "mapme.vegetation"), 
                      dx = dx, 
                      dy = dy, 
                      dt = dt, 
                      srs = srs, 
                      bbox = bbox, 
                      after = after, 
                      before = before, 
                      aggregation = aggregation, 
                      resampling = resampling, 
                      index = index, 
                      label = label, 
                      outdir = file.path(rundir, "ndvi"), 
                      overwrite = TRUE,
                      mask_layer = mask_layer, 
                      mask_values = mask_values,
                      mask_invert = mask_invert,
                      creation_options = list("COMPRESS" = "DEFLATE"))
```

Let's inspect the returned output. It is a list with three objects. The first 
object is called `files` and contains the path to the files. The second object 
is called `band_order` and indicates the names and order of the bands in the 
output files. This can be very important when multiple indices are calculated
at once because `gdalcubes` will return the bands in alphabetical order.
The last object is called `time` and indicates the timestamp for each of the 
output files. Together this information can be used for subsequent processing.

```{r}
str(raw_ndvi)
```

Let's visualize the NDVI raster files.

```{r}
ndvi_ras = rast(raw_ndvi$files)
plot(ndvi_ras)

```

We see that we have some missing values during the first 14 day period of May 
as well as more extensive missing values for the period starting of May, 29th.
Note that in the original resolution it is likely that there are more missing values.
In order to be able to ship the example data sets with the package, the spatial
resolution has been resampled to 100 meters, which is a reduction in spatial
resolution of factor 10 for the band 4 and 8 that are used for the NDVI calculation.
However, due to cloud cover, missing observations are actually quite frequent.
The `mapme.vegetation` package offers functionality for to apply interpolation of
missing values and smoothing of the resulting time series. 

## Smoothing a Time Series

The default functionality is quite basic, applying linear interpolation and 
applying a Savitzkiy-Golay filter. In case you need a more sophisticated gap-filling
strategy you can define your own function. The function receives the time series 
of each pixel and is expected to return the same number of pixels. Let's 
consider the smoothing with the basic approach. We will set the chunking option to 
simultaneously process all 7 time-steps of our spatio-temporal extent. This is 
important so that the complete time series is available.

```{r}

fill_ndvi = smooth_cube(files = raw_ndvi$files,
                        bands = raw_ndvi$band_order, 
                        times = raw_ndvi$time,
                        dx = dx, 
                        dy = dy, 
                        dt = dt, 
                        srs = srs, 
                        after = after, 
                        before = before, 
                        bbox = bbox, 
                        aggregation = aggregation, 
                        resampling = resampling, 
                        outdir = file.path(rundir, "filled"), 
                        label = label,
                        overwrite = TRUE,
                        chunking = c(7,100,100),
                        creation_options = list("COMPRESS" = "DEFLATE"))
```

Let's visualize the resulting gap filled and smoothed raster files. We see that the 
missing values have been filled. Whether or not the default smoothing function 
is suitable for your use case is a question you have to answer for yourself.
In the example section of `mapme.vegetation::smooth_cube` you will find the 
default. You can exchange that function by specifying a custom function with the
`smoothing_function` argument. 

```{r}
ndvi_ras2 = rast(fill_ndvi$files)
plot(ndvi_ras2)
```


```{r rm-rundir, echo = FALSE, message=FALSE, show = FALSE}
unlink(rundir, recursive = T)
```
