---
title: "Installation"
date: "Last modified: `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This sections describes how to install the `{mapme.vegetation}` package. 
Because the package handles some spatial data, we assume that you have
a working installation of [GDAL](https://gdal.org/index.html) and [PROJ](https://proj.org/) 
on your system. Please refer to the respective installation guides to learn about 
how to install these programs on your OS.
Additionally, the cloud masking routine works by using the [GRASS GIS](https://grass.osgeo.org/) software.
Be referred to the indicated websites to learn how to install this programm.
While the package supports sequentially downloading Sentinel-2 images from AWS,
this could take a long time when there are many assets to download. To speed
up this process by parallel downloading several files we support the [aria2](https://aria2.github.io/) 
downloader. We expect that the path to the executable binaries to both a 
working GRASS GIS installation and the aria2 downloader is known to the user
and specified to the respective function calls.

Windows users additionally will have to install the [rtools](https://cran.r-project.org/bin/windows/Rtools) 
software for a successful installation of the `{mapme.vegetation}` package from source. 
 
Installing the `{mapme.vegetation}` package is then achieved trough the following
function call: 


```{r installation, eval = F}
remotes::install_github("mapme-initiative/mapme.vegetation")
```

The installation process will install potential missing R dependencies.
After the successful installation, you are ready to use the package's 
functionality. Simply load the package for your R session with the library
call:

```{r setup, val = F}
library(mapme.vegetation)
```

