---
title: "Terminology"
date: "Last modified: `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Terminology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

Here we present a quick introduction and the most important terminology used throughout
this tutorial:


-  Surface Reflectance: The surface reflectance is is measured by satellites in different
bands across the electromagnetic spectrum. Depending on the characteristics of a surface
the spectral response for different bands changes. This information can be harnessed,
e.g. to differentiate between different types of surfaces.

- Vegetation Index: A vegetation index usually is calculated by the inclusion of one
or more surface reflectance bands from a remote sensing data. For example, the 
most well-known index, the normalized difference vegetation index (NDVI) considers
the difference between the red (R) and near-infrared (NIR) bands following:
$\frac{(NIR-R)}{(NIR + R)}$

- Sparse/Dense Time Series: A timeseries does have pre-defined time interval when observations
are repeated. In a sparse timeseries a number of these regular interval observations
are missing, e.g. in the case of Sentinel-2 data through cloud cover. By aggregating 
to a larger time-interval between observations we can derive a dense time series where
each value is a composite value by the more fine-grained original observational data.
To generate such a dense timeseries is the purpose of the pre-processing functionality 
of `{mapme.vegetation}`

-  Trend Analysis: This is the process of analyzing the behavior of a measurement
through time. For example, in terms of VIs, we could be interested if the
yearly maximum observed VI value increases from one year to another. In this 
case we would fit a linear trend model with the maximum VI as a dependent and time
as an independent variable. 

-  Zonal Statistics: Satellite data usually comes in a raster format, that is 
regularly spaces grid cells covering the earth's surface. Sometimes we might
not be interested in a single one of these cells, but we are instead interested
in an area covering multiple cells. An example would be an agricultural field potentially
covering some hundreds of grid cells. To get information for the object of interest, 
that is the whole field, we need to spatially aggregate the information of the grid
cells. This is achieved through zonal statistics, e.g.the calculation of the 
minimum, the average, and the maximum value of all pixels within a specific zone,
i.e. a field in the example above. 

