---
title: "Calculate Trends"
date: "Last modified: `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculate Trends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This tool measures the rate of change in vegetation productivity over time, based 
on vegetation indices from multi-temporal satellite observations. It calculates a 
linear regression at the pixel level to measure the trajectory of change in vegetation 
productivity over time for the period under analysis. A Mann-Kendall non-parametric 
significance test is then applied. This allows the user to consider, i.e. to map 
only significant changes that show a certain p-value (e.g. ≤ 0.05). Positive 
significant slopes would indicate potential improvement (e.g. vegetation gain), 
while negative significant slopes could be indicative of degradation (vegetation loss). 
Please note that it is always recommended to cross check and validate the results 
from this assessment with in-situ measurements, if available.

In the following we will show how to calculate the trend in a time-series of vegetation 
indices and how to compare two different epochs, e.g. pre- and post-intervention period
As a prerequisite we assume that you have already calculated a dense time-series
of the index of your choice for the relevant timeframes. The example we are 
presenting here might not make a lot of sense of a analytic level, but it is 
only meant to introduce the functionality. 
Let's get started by reading in our dense NDVI time-series and separate it into
a rather random pre- and post-intervention period

```{r setup}
library(mapme.vegetation)
library(stringr)
library(sf)
library(terra)

ndvi_files = s2_files = list.files(system.file("extdata/filled", package = "mapme.vegetation"), ".tif", full.names = T)
bands = "NDVI"
times = str_sub(basename(ndvi_files), -14, -5)

aoi = st_read(system.file("extdata", "exp_region_wgs.gpkg", package = "mapme.vegetation"), quiet = T)
aoi_utm = st_transform(aoi, st_crs(32638))
aoi_utm$id = 1
bbox = st_bbox(aoi_utm)

pre = ndvi_files[1:4]
post = ndvi_files[5:7]
times_pre = times[1:4]
times_post = times[5:7]

```

We will have to process each of these periods separately since the function
will try to fit a trend for all the available time-steps at once. By default
the trend will be estimated using the `trend::mk.test()` function which fits 
a simple linear trend. Two values are returned, that is the estimate of the 
slope of the linear trend as well as the p.value. Via the argument `filter_p`
we can exclude pixels below a certain threshold. However, we will not apply
this in this tutorial as we do not expect to find significant trends in this
rather random time periods.

```{r warning=FALSE, message=F}
trend_pre = trend_cube(files = pre, 
                       bands = bands, 
                       times = times_pre, 
                       dx = 200, 
                       dy = 200,
                       dt = "P14D",
                       srs = "EPSG:32638",
                       after = times_pre[1],
                       before = times_pre[4],
                       bbox = bbox,
                       aggregation = "max",
                       resampling = "cubic",
                       chunking = c(4, 256, 256),
                       outdir = "../inst/extdata/trend/",
                       label = "pre_")

trend_post = trend_cube(files = post, 
                        bands = bands, 
                        times = times_post, 
                        dx = 200, 
                        dy = 200,
                        dt = "P14D",
                        srs = "EPSG:32638",
                        after = times_post[1],
                        before = times_post[3],
                        bbox = bbox, 
                        aggregation = "max",
                        resampling = "cubic",
                        chunking = c(3, 256, 256),
                        outdir = "../inst/extdata/trend/",
                        label = "post_")
```

Let's visualize the trend results for both the pre- and post intervention period.
Note that the first layer returned by the default trend function is the slope of
the linear trend while the second layer defaults to the p-value.

```{r}
pre = rast(system.file("extdata/trend/pre_2017-05-01.tif", package = "mapme.vegetation"))
names(pre) = c("PRE:slope", "PRE:p.value")

post = rast(system.file("extdata/trend/post_2017-06-26.tif", package = "mapme.vegetation"))
names(post) = c("POST:slope", "POST:p.value")


comp = c(pre,post)
plot(comp)
```

Having calculated the NDVI slope for both time periods, we now might be interested
in learning the differences between these two. The `{mapme.vegetation}` package
comes with a function called `compare_epochs()` for just that. It allows the
pixel-wise comparison between two rasters. Multiple layers are understood as a
time dimension and a suitable temporal aggregation method can be chosen before
the comparison is conducted. User's are expected to specify one simple comparison
operator (e.g. + / - / : / * ) and the order of the comparision (which epoch comes
first). In our example, let's calculate the fraction of the slope we observe
during the post period in relation to the pre period. Values below one will indicate
pixels which show a smaller slope in the post intervention period, while pixels with
values greater one will indicate areas with greater slopes compared to the 
pre-intervention period. The function excepts either character strings pointing
to raster files or spatRaster objects.

```{r}
(ep_comp = compare_epochs(epoch1_files = pre$`PRE:slope` , 
                          epoch2_files = post$`POST:slope`, 
                          order = c(2,1), 
                          stat = "/"))
plot(ep_comp)

```

We do not see much but why? Let's inspect the value range.

```{r}
boxplot(values(ep_comp))
```

That is interesting! There is a single pixels that shows a difference in NDVI between
the pre- and post-intervention period of up to 80,000% (note that you have to multiply
with 100 to get percent). Because of that very large value range we don't see much when
we plot the data. Anyway, a difference that large is highly implausible. The main
reason is that our comparison between pre- and post-intervention does not make any
sense. The split into the two different periods as well as the trend calculation
based only on 3 or 4 data points is not very sensual. However, we showcased the
functionality of the `{mapme.vegetation}` package to calculate and compare trends
between different epochs as these can be used in real world analysis of project
evaluations.
