---
title: "Workflow"
date: "Last modified: `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 5
)
```

In the following, the most important steps of the processing workflow 
of `{mapme.vegetation}` are described; the figure guides the reader through these
descriptions. The overall goal here is to download the matching Sentinel-2 tiles
a spatio-temporal extent of interest and process the downloaded data to 
derive a dense time series of surface reflectance bands and vegetation indices. 
On the other end of this process, there are optional additional analysis steps
such as the calculation of pixel-wise trends or the extraction of zonal statistics. 

-  download only relevant bands for a given spatio-temporal extent of Sentinel-2 L2A data
-  prepare a custom cloud mask based on the sen2cor Scene Cover Classification (SCL)
-  calculate indices and extract surface reflectance bands and aggregate values to a regular sparse timeseries
-  apply an interpolation and smoothing strategy to fill in missing data, reduce noise, and create a dense timeseires
-  optionally, calculate pixel-wise trends through the time dimension
-  optionally, extract zonal statistics of areas of interest, e.g. for diff-in-diff analysis
-  optionally, extract pixel time-series, e.g. for later usage with the <code>{[mapme.classification](https://mapme-initiative.github.io/mapme.classification/)}</code>


```{r wflow, echo = FALSE}
library(DiagrammeR)

mermaid("
  graph TB

    classDef default fill:#c2dddd,stroke-width:2px,stroke:#034858;
 
    A(Query & Download)
    B(Apply Cloud Mask)
    C(Calculate VIs and SR)
    D(Dense Time Series)
    E(Trends)
    F(Zonal Statistics)
    G(Pixel Extraction)
    A-->B
    B-->C
    C-->D
    D-->E
    D-->F
    D-->G

")

```

