---
title: "Extract Zonal Statistics"
date: "Last modified: `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
%\VignetteIndexEntry{Extract Zonal Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  ---
  
  ```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```

```{r setup, warning=F}
library(mapme.vegetation)
library(sf)
library(stringr)
```


This vignette showcases different ways to extract zonal and pixelwise statistics
from a time series processed with the `mapme.vegetation` package. We assume
that we have processed S2 data to a dense time-series following vignette xx.
Let's read in the files and our AOI.

```{r}
ndvi_files = s2_files = list.files(system.file("extdata/filled", package = "mapme.vegetation"), ".tif", full.names = T)
bands = "NDVI"
times = str_sub(basename(ndvi_files), -14, -5)

aoi = st_read(system.file("extdata", "exp_region_wgs.gpkg", package = "mapme.vegetation"), quiet = T)
aoi_utm = st_transform(aoi, st_crs(32638))
aoi_utm$id = 1
bbox = st_bbox(aoi_utm)
```

As an example we will extract zonal statistics for the complete AOI. This function
relies on `gdalcubes` which is why we need information of the spatio-temporal extent
of the input files. These are the same as in the other vignettes. With this
function we get information averaged over all pixels within a polygon for each 
time step. This information is useful for statistical analysis and might also 
be suitable for object based classification or regression problems which require
information through the time axis. Specific metrics can be calculated. In case
`zonalstat = "all"`, the count of pixels within the polygon, max, min, mean, median,
prod, sd, sum, and var are returned. It is important to declare a column as the
primary identification column with the argument `idcol`. Note that the returned
object will be in a time-long format. That also means that the geometry column will be
repeated. Because of this, the resulting object might be quite large. You can
specify a filepath with the `outpath` argument to write the object to a GeoPackage (.gpkg).

```{r}
zs_aoi = extract_zonalstats(aoi = aoi_utm, 
                            idcol = "id", 
                            files = ndvi_files, 
                            times = times, 
                            bands = bands, 
                            bbox = bbox, 
                            zonalstat = "all", 
                            after = "2017-05-01", 
                            before = "2017-07-31", 
                            srs = "EPSG:32638", 
                            dx = 200, 
                            dy = 200, 
                            dt = "P14D", 
                            aggregation = "max", 
                            resampling = "near")
head(zs_aoi)
```

On other occasions pixel-wise information for polygons might be of interest, e.g.
when the goal of your analysis is the pixel-wise classification of a landscape.
For this kind of extraction we rely on the `{terra}` package. For the resulting 
data.frame object to receive the right column names, however, it is important
to hand over the band names and the associated time-stamp for each file.

```{r}
ps_aoi = extract_pixels(files = ndvi_files,
                        bands = bands,
                        times = times, 
                        aoi = aoi_utm)
head(ps_aoi)
```

The resulting object is **not** an sf object. The spatially explicit information
is list. However, the column `ID` indicates to to which feature in the `aoi` object
the pixel information belongs in the order of rows of the input object. The data.frame
object has a wide format. Each row represents a single pixel. The columns identify
the band name as well as the time-stamp.
